---
title: "Where Does the Minimum Wage Go Furthest?"
date: 2026-09-20
categories:
  - opinion
  - labor market
  - cost of living
  
format:
  html:
    toc: false
execute:
  echo: false
  warning: false
  message: false
---

<style>
main.content {
  font-family: Arial, Helvetica, sans-serif;
  font-size: 17px;
  line-height: 1.8;
}
h1.title {
  font-size: 2.2rem;
  line-height: 1.25;
}
main.content h2 {
  font-size: 1.55rem;
  margin-top: 2rem;
  margin-bottom: 0.8rem;
}
main.content p {
  margin-bottom: 1.1rem;
}
.quarto-figure {
  margin-top: 1.5rem;
  margin-bottom: 1.5rem;
}
figcaption {
  font-size: 0.9rem;
  line-height: 1.5;
}
</style>

```{r}
#| label: setup

# Find the folder containing this index.qmd.
input_file <- knitr::current_input(dir = TRUE)

post_dir <- dirname(normalizePath(
  input_file,
  winslash = "/",
  mustWork = TRUE
))

# Check that the scraping script has been saved correctly.
scraper_file <- file.path(post_dir, "code", "01_scrape.R")

if (!file.exists(scraper_file)) {
  stop(
    "Cannot find the scraping script. Save 01_scrape.R in: ",
    file.path(post_dir, "code"),
    call. = FALSE
  )
}

# Check required packages.
required_packages <- c("rvest", "xml2", "httr", "stringr")

missing_packages <- required_packages[
  !vapply(
    required_packages,
    requireNamespace,
    quietly = TRUE,
    FUN.VALUE = logical(1)
  )
]

if (length(missing_packages)) {
  stop(
    "Run this in the RStudio Console first:\n",
    'install.packages(c("',
    paste(missing_packages, collapse = '", "'),
    '"))',
    call. = FALSE
  )
}

# Load and run the actual two-website scraper.
source(scraper_file, local = TRUE)

# First run downloads the websites.
# Later renders reuse the saved HTML.
# Set TRUE only when intentionally refreshing the source data.
wages <- scrape_blog2(
  output_dir = post_dir,
  refresh = FALSE
)

money <- function(x) {
  sprintf("$%.2f", x)
}

value <- function(state_name, column) {
  result <- wages[
    wages$state == state_name,
    column
  ]

  if (length(result) != 1L || is.na(result)) {
    stop(
      "Cannot obtain a single value for ",
      state_name, ": ", column,
      call. = FALSE
    )
  }

  result[[1]]
}

english_list <- function(x) {
  x <- sort(unique(x))

  if (!length(x)) return("None")
  if (length(x) == 1L) return(x)
  if (length(x) == 2L) return(paste(x, collapse = " and "))

  paste0(
    paste(head(x, -1), collapse = ", "),
    ", and ",
    tail(x, 1)
  )
}

top10 <- head(wages, 10)
sample_n <- nrow(wages)

wage_date <- unique(wages$effective_date)
cost_period <- unique(wages$cost_period)

stopifnot(
  length(wage_date) == 1L,
  length(cost_period) == 1L
)

all_data <- read.csv(
  file.path(post_dir, "data/cleaned/all_jurisdictions.csv"),
  stringsAsFactors = FALSE
)

range_data <- read.csv(
  file.path(post_dir, "data/cleaned/multi_rate_states.csv"),
  stringsAsFactors = FALSE
)

matched_n <- nrow(all_data)
excluded_names <- english_list(range_data$state)

nominal_top <- wages[
  which.max(wages$wage),
  ,
  drop = FALSE
]

ca_equivalent <- value("Missouri", "adjusted") *
  value("California", "cost_index") / 100

dol_metadata <- read.csv(
  file.path(post_dir, "data/raw/dol/source_metadata.csv"),
  stringsAsFactors = FALSE
)

meric_metadata <- read.csv(
  file.path(post_dir, "data/raw/meric/source_metadata.csv"),
  stringsAsFactors = FALSE
)

dol_retrieved <- substr(dol_metadata$retrieved_utc[1], 1, 10)
meric_retrieved <- substr(meric_metadata$retrieved_utc[1], 1, 10)

# Save the figures generated by the article.
knitr::opts_chunk$set(
  fig.path = "results/figures/",
  dev = "png",
  dpi = 160
)

# Create the project README alongside index.qmd.
readme_lines <- c(
  "# Blog 2: Where Does the Minimum Wage Go Furthest?",
  "",
  "## Research question",
  "How does adjusting minimum wages for state-level living costs change the comparison across U.S. jurisdictions?",
  "",
  "## Sources",
  "- U.S. Department of Labor: https://www.dol.gov/agencies/whd/mw-consolidated",
  "- MERIC: https://meric.mo.gov/data/cost-living-data-series",
  "",
  paste0("DOL wage schedule: ", wage_date, "."),
  paste0("MERIC cost period: ", cost_period, "."),
  paste0("DOL retrieval date (UTC): ", dol_retrieved, "."),
  paste0("MERIC retrieval date (UTC): ", meric_retrieved, "."),
  "",
  "## Files",
  "- code/01_scrape.R: downloads both webpages, extracts tables with rvest, cleans and joins observations, and saves CSV outputs.",
  "- data/raw/dol/: original DOL HTML, extracted cells, and retrieval metadata.",
  "- data/raw/meric/: original MERIC HTML, extracted table, and retrieval metadata.",
  "- data/cleaned/: merged observations, the single-rate analysis sample, and multiple-rate states.",
  "- results/figures/: figures generated when index.qmd is rendered.",
  "- results/table/: rankings, the top ten, the multiple-rate appendix, and validation counts.",
  "- index.qmd: article text, calculations, figures, and tables.",
  "- README.md: project description and reproduction instructions.",
  "",
  "## Reproduce",
  "Install R, RStudio, and Quarto.",
  "",
  "Install the required R packages:",
  "",
  "```r",
  'install.packages(c("rvest", "xml2", "httr", "stringr", "knitr", "rmarkdown"))',
  "```",
  "",
  "Open index.qmd in RStudio and click Render.",
  "The first run downloads both websites. Subsequent runs parse the saved HTML.",
  "To request current webpages, change refresh = FALSE to refresh = TRUE in the setup chunk, render once, and then change it back to FALSE.",
  "From a terminal in this post folder, the equivalent rendering command is: quarto render index.qmd",
  "",
  "## Method",
  "Cost-adjusted hourly wage = nominal hourly minimum wage / (cost index / 100).",
  "The wage scope is general non-tipped employment covered by the federal FLSA.",
  "States without a higher applicable state floor use the federal floor.",
  "Superscript footnote markers are removed before wage values are parsed.",
  "Multiple general rates are retained in a separate appendix rather than reduced to an arbitrary single value.",
  paste0("Matched jurisdictions: ", matched_n, "."),
  paste0("Main analysis sample: ", sample_n, "."),
  paste0("Multiple-rate states excluded from the main comparison: ", excluded_names, "."),
  "",
  "## Limitations",
  "The wage schedule and cost index cover different periods.",
  "State averages conceal local variation.",
  "The adjustment omits taxes, benefits, household composition, and job availability.",
  "The results describe a purchasing-power proxy and do not identify causal effects.",
  "Refreshing requires internet access; changed website layouts or access rules may require updating the scraper.",
  "",
  "## Source notice",
  "The article includes the State of Missouri source-data notice."
)

writeLines(
  readme_lines,
  file.path(post_dir, "README.md"),
  useBytes = TRUE
)
```

## A Higher Wage, but a Better Deal?

If two jobs offer different hourly wages, the higher offer initially looks better. But rent, groceries, and other expenses vary across places. For workers considering jobs near the wage floor, a useful question is: **where does the minimum wage buy more after accounting for local prices?** I combine two public websites to build a simple comparison. The goal is a practical screening tool for comparing locations, rather than a claim that one state is best for every worker.

## The Data Behind the Comparison

This analysis combines the [U.S. Department of Labor’s minimum-wage table](https://www.dol.gov/agencies/whd/mw-consolidated), effective `r wage_date`, with [MERIC’s cost-of-living index](https://meric.mo.gov/data/cost-living-data-series) for `r cost_period`. MERIC averages indices from participating cities and metropolitan areas within each state, with the national benchmark set at 100. Because the wage schedule and cost index cover different periods, the published cost figures serve as a price benchmark rather than a precise measure of prices when the wage rates took effect.

After cleaning, the two datasets were matched across all 50 states and Washington, DC. For employment covered by the federal Fair Labor Standards Act, states without a higher state minimum use the federal minimum wage—not zero. States listing multiple general rates are retained separately with all listed rates preserved, rather than assigned a single value. The main comparison therefore includes **`r sample_n` jurisdictions**. Special local, industry-specific, youth, and tipped-worker wage rates are outside the scope of this analysis.

The calculation is:

$$
\text{Cost-adjusted hourly wage}
=
\frac{\text{Hourly minimum wage}}
{\text{Cost-of-living index}/100}
$$

An index of 120 reduces the nominal wage to its equivalent at the national price benchmark; an index of 90 increases it. This measure is a purchasing-power proxy, not a new legal wage, take-home pay, or a complete living-wage estimate.

## Prices Change the Ranking

`r wages$state[1]` leads the main sample after adjustment at approximately **`r money(wages$adjusted[1])` per hour**. `r nominal_top$state` has the highest nominal floor in the sample, `r money(nominal_top$wage)`, equivalent to `r money(nominal_top$adjusted)` after adjustment. California's `r money(value("California", "wage"))` floor becomes about `r money(value("California", "adjusted"))`. Missouri's `r money(value("Missouri", "wage"))` becomes `r money(value("Missouri", "adjusted"))`. Massachusetts has a `r money(value("Massachusetts", "wage"))` floor, equivalent to `r money(value("Massachusetts", "adjusted"))`. The comparison shows why a wage's dollar value alone is not enough to assess purchasing power.

```{r}
#| label: fig-top-ten
#| fig-cap: "The ten highest cost-adjusted minimum wages in the main sample. Connected points compare nominal and adjusted hourly amounts."
#| fig-width: 8
#| fig-height: 5.5
#| out-width: "100%"
#| fig-align: center

local({
  old <- par(no.readonly = TRUE)
  on.exit(par(old))

  d <- top10[nrow(top10):1, ]
  y <- seq_len(nrow(d))

  par(
    mar = c(4.5, 9, 1.5, 1),
    las = 1
  )

  plot(
    NA,
    xlim = range(c(d$wage, d$adjusted)) + c(-0.6, 0.6),
    ylim = c(0.5, nrow(d) + 1.8),
    xlab = "Hourly amount (US dollars)",
    ylab = "",
    yaxt = "n",
    bty = "n"
  )

  abline(
    h = y,
    col = "#EEEEEE"
  )

  segments(
    d$wage, y,
    d$adjusted, y,
    col = "#BBBBBB",
    lwd = 2
  )

  points(
    d$wage, y,
    pch = 16,
    col = "#777777",
    cex = 1.2
  )

  points(
    d$adjusted, y,
    pch = 16,
    col = "#236B8E",
    cex = 1.2
  )

  axis(
    2,
    at = y,
    labels = d$state,
    tick = FALSE,
    cex.axis = 0.85
  )

  legend(
    "topright",
    legend = c("Nominal", "Cost-adjusted"),
    col = c("#777777", "#236B8E"),
    pch = 16,
    horiz = TRUE,
    bty = "n",
    cex = 0.85
  )
})
```

The full-sample scatterplot shows how the adjustment works beyond a few examples. Points above the diagonal have cost indices below 100; those below it have higher indices. The descriptive pattern does not establish that wage laws cause prices, employment, or migration. It shows why comparing dollar amounts alone can miss an important part of a worker's budget.

```{r}
#| label: fig-all-states
#| fig-cap: "Nominal versus cost-adjusted minimum wages in the main sample. The dashed line indicates equal nominal and adjusted values."
#| fig-width: 8
#| fig-height: 5.5
#| out-width: "100%"
#| fig-align: center

local({
  old <- par(no.readonly = TRUE)
  on.exit(par(old))

  par(
    mar = c(4.5, 4.5, 1, 1),
    las = 1
  )

  limits <- range(
    c(wages$wage, wages$adjusted),
    finite = TRUE
  ) + c(-1, 1.5)

  plot(
    wages$wage,
    wages$adjusted,
    xlim = limits,
    ylim = limits,
    xlab = "Nominal hourly minimum wage (US dollars)",
    ylab = "Cost-adjusted hourly wage (US dollars)",
    pch = 16,
    col = "#236B8E99",
    cex = 1.05,
    bty = "l"
  )

  abline(
    a = 0,
    b = 1,
    lty = 2,
    col = "#888888"
  )

  selected_states <- c(
    "Missouri",
    "California",
    "Massachusetts",
    "District of Columbia"
  )

  selected <- wages[
    match(selected_states, wages$state),
    ,
    drop = FALSE
  ]

  points(
    selected$wage,
    selected$adjusted,
    pch = 16,
    col = "#173F56",
    cex = 1.2
  )

  text(
    selected$wage,
    selected$adjusted,
    labels = c("MO", "CA", "MA", "DC"),
    pos = c(3, 4, 1, 3),
    cex = 0.85,
    offset = 0.6
  )
})
```

## What Should a Job Seeker Do?

A worker comparing otherwise similar minimum-wage opportunities could use the adjusted ranking to shortlist locations, then check actual offers and local housing costs. Matching Missouri's minimum-wage purchasing-power proxy in California would require approximately **`r money(ca_equivalent)` per hour** at these indices. This is a benchmark calculation, not a forecast or salary recommendation. The practical lesson is to compare **the actual offered wage relative to relevant local costs**, then consider hours, benefits, job availability, and moving expenses.

## Important Limitations

State averages hide substantial differences between cities, and MERIC's participating areas need not represent every worker's spending. The proxy omits taxes, benefits, job-finding chances, and personal household needs. Wage coverage also varies, and the excluded multi-rate states may suit particular workers better than the main ranking suggests. These limitations make the results a starting point for research, not a complete relocation recommendation.

## Top Ten: The Numbers

```{r}
#| label: table-top-ten

table_out <- data.frame(
  State = top10$state,
  `Minimum wage` = money(top10$wage),
  `Cost index` = sprintf("%.1f", top10$cost_index),
  `Adjusted wage` = money(top10$adjusted),
  check.names = FALSE
)

knitr::kable(
  table_out,
  align = c("l", "r", "r", "r"),
  caption = paste0(
    "Highest adjusted hourly amounts in the ",
    sample_n,
    "-jurisdiction sample."
  )
)
```

## Appendix: States with Multiple Rates

The states excluded from the single-rate comparison are: **`r excluded_names`**. The endpoints below apply the same state cost index to the lowest and highest listed general wage rates. They are scenarios, not confidence intervals; geographic and employer-specific coverage still matter.

```{r}
#| label: table-multiple-rates

if (nrow(range_data) > 0) {

  range_out <- data.frame(
    State = range_data$state,
    `Nominal range` = paste0(
      money(range_data$wage_low),
      " to ",
      money(range_data$wage_high)
    ),
    `Adjusted range` = paste0(
      money(range_data$adjusted_low),
      " to ",
      money(range_data$adjusted_high)
    ),
    check.names = FALSE
  )

  knitr::kable(
    range_out,
    align = c("l", "r", "r"),
    caption = "Multiple-rate states retained outside the main ranking."
  )
}
```

## Data Sources

- [U.S. Department of Labor: Consolidated State Minimum Wage Update Table](https://www.dol.gov/agencies/whd/mw-consolidated). Effective `r wage_date`. Retrieved `r dol_retrieved` (UTC).
- [MERIC: Cost of Living Data Series](https://meric.mo.gov/data/cost-living-data-series). `r cost_period`; underlying survey by C2ER. Retrieved `r meric_retrieved` (UTC).

<details>
<summary>Missouri source-data notice</summary>

The data made available here has been modified for use from its original source, which is the State of Missouri. THE STATE OF MISSOURI MAKES NO REPRESENTATIONS OR WARRANTY AS TO THE COMPLETENESS, ACCURACY, TIMELINESS, OR CONTENT OF ANY DATA MADE AVAILABLE THROUGH THIS SITE. THE STATE OF MISSOURI EXPRESSLY DISCLAIMS ALL WARRANTIES, WHETHER EXPRESS OR IMPLIED, INCLUDING ANY IMPLIED WARRANTIES OF MERCHANTABILITY, OR FITNESS FOR A PARTICULAR PURPOSE.

The data is subject to change as modifications and updates are complete. It is understood that the information contained in the Web feed is being used at one's own risk.

[State of Missouri Data Policy](https://mo.gov/data-policy)

</details>